---
title: "Multinomial Logit Model"
author: "Siddharth Bam"
date: today
---


This assignment expores two methods for estimating the MNL model: (1) via Maximum Likelihood, and (2) via a Bayesian approach using a Metropolis-Hastings MCMC algorithm. 


## 1. Likelihood for the Multi-nomial Logit (MNL) Model

Suppose we have $i=1,\ldots,n$ consumers who each select exactly one product $j$ from a set of $J$ products. The outcome variable is the identity of the product chosen $y_i \in \{1, \ldots, J\}$ or equivalently a vector of $J-1$ zeros and $1$ one, where the $1$ indicates the selected product. For example, if the third product was chosen out of 3 products, then either $y=3$ or $y=(0,0,1)$ depending on how we want to represent it. Suppose also that we have a vector of data on each product $x_j$ (eg, brand, price, etc.). 

We model the consumer's decision as the selection of the product that provides the most utility, and we'll specify the utility function as a linear function of the product characteristics:

$$ U_{ij} = x_j'\beta + \epsilon_{ij} $$

where $\epsilon_{ij}$ is an i.i.d. extreme value error term. 

The choice of the i.i.d. extreme value error term leads to a closed-form expression for the probability that consumer $i$ chooses product $j$:

$$ \mathbb{P}_i(j) = \frac{e^{x_j'\beta}}{\sum_{k=1}^Je^{x_k'\beta}} $$

For example, if there are 3 products, the probability that consumer $i$ chooses product 3 is:

$$ \mathbb{P}_i(3) = \frac{e^{x_3'\beta}}{e^{x_1'\beta} + e^{x_2'\beta} + e^{x_3'\beta}} $$

A clever way to write the individual likelihood function for consumer $i$ is the product of the $J$ probabilities, each raised to the power of an indicator variable ($\delta_{ij}$) that indicates the chosen product:

$$ L_i(\beta) = \prod_{j=1}^J \mathbb{P}_i(j)^{\delta_{ij}} = \mathbb{P}_i(1)^{\delta_{i1}} \times \ldots \times \mathbb{P}_i(J)^{\delta_{iJ}}$$

Notice that if the consumer selected product $j=3$, then $\delta_{i3}=1$ while $\delta_{i1}=\delta_{i2}=0$ and the likelihood is:

$$ L_i(\beta) = \mathbb{P}_i(1)^0 \times \mathbb{P}_i(2)^0 \times \mathbb{P}_i(3)^1 = \mathbb{P}_i(3) = \frac{e^{x_3'\beta}}{\sum_{k=1}^3e^{x_k'\beta}} $$

The joint likelihood (across all consumers) is the product of the $n$ individual likelihoods:

$$ L_n(\beta) = \prod_{i=1}^n L_i(\beta) = \prod_{i=1}^n \prod_{j=1}^J \mathbb{P}_i(j)^{\delta_{ij}} $$

And the joint log-likelihood function is:

$$ \ell_n(\beta) = \sum_{i=1}^n \sum_{j=1}^J \delta_{ij} \log(\mathbb{P}_i(j)) $$



## 2. Simulate Conjoint Data

We will simulate data from a conjoint experiment about video content streaming services. We elect to simulate 100 respondents, each completing 10 choice tasks, where they choose from three alternatives per task. For simplicity, there is not a "no choice" option; each simulated respondent must select one of the 3 alternatives. 

Each alternative is a hypothetical streaming offer consistent of three attributes: (1) brand is either Netflix, Amazon Prime, or Hulu; (2) ads can either be part of the experience, or it can be ad-free, and (3) price per month ranges from \$4 to \$32 in increments of \$4.

The part-worths (ie, preference weights or beta parameters) for the attribute levels will be 1.0 for Netflix, 0.5 for Amazon Prime (with 0 for Hulu as the reference brand); -0.8 for included adverstisements (0 for ad-free); and -0.1*price so that utility to consumer $i$ for hypothethical streaming service $j$ is 

$$
u_{ij} = (1 \times Netflix_j) + (0.5 \times Prime_j) + (-0.8*Ads_j) - 0.1\times Price_j + \varepsilon_{ij}
$$

where the variables are binary indicators and $\varepsilon$ is Type 1 Extreme Value (ie, Gumble) distributed.

The following code provides the simulation of the conjoint data.

:::: {.callout-note collapse="true"}
```{r}
# set seed for reproducibility
set.seed(123)

# define attributes
brand <- c("N", "P", "H") # Netflix, Prime, Hulu
ad <- c("Yes", "No")
price <- seq(8, 32, by=4)

# generate all possible profiles
profiles <- expand.grid(
    brand = brand,
    ad = ad,
    price = price
)
m <- nrow(profiles)

# assign part-worth utilities (true parameters)
b_util <- c(N = 1.0, P = 0.5, H = 0)
a_util <- c(Yes = -0.8, No = 0.0)
p_util <- function(p) -0.1 * p

# number of respondents, choice tasks, and alternatives per task
n_peeps <- 100
n_tasks <- 10
n_alts <- 3

# function to simulate one respondent’s data
sim_one <- function(id) {
  
    datlist <- list()
    
    # loop over choice tasks
    for (t in 1:n_tasks) {
        
        # randomly sample 3 alts (better practice would be to use a design)
        dat <- cbind(resp=id, task=t, profiles[sample(m, size=n_alts), ])
        
        # compute deterministic portion of utility
        dat$v <- b_util[dat$brand] + a_util[dat$ad] + p_util(dat$price) |> round(10)
        
        # add Gumbel noise (Type I extreme value)
        dat$e <- -log(-log(runif(n_alts)))
        dat$u <- dat$v + dat$e
        
        # identify chosen alternative
        dat$choice <- as.integer(dat$u == max(dat$u))
        
        # store task
        datlist[[t]] <- dat
    }
    
    # combine all tasks for one respondent
    do.call(rbind, datlist)
}

# simulate data for all respondents
conjoint_data <- do.call(rbind, lapply(1:n_peeps, sim_one))

# remove values unobservable to the researcher
conjoint_data <- conjoint_data[ , c("resp", "task", "brand", "ad", "price", "choice")]

# clean up
rm(list=setdiff(ls(), "conjoint_data"))
```
::::



## 3. Preparing the Data for Estimation

The "hard part" of the MNL likelihood function is organizing the data, as we need to keep track of 3 dimensions (consumer $i$, covariate $k$, and product $j$) instead of the typical 2 dimensions for cross-sectional regression models (consumer $i$ and covariate $k$). The fact that each task for each respondent has the same number of alternatives (3) helps.  In addition, we need to convert the categorical variables for brand and ads into binary variables.

_todo: reshape and prep the data_
```{r}
# Load required package
library(dplyr)

# Step 1: One-hot encode 'brand' and 'ad'
mnl_data <- conjoint_data %>%
  mutate(
    Netflix = ifelse(brand == "N", 1, 0),
    Prime   = ifelse(brand == "P", 1, 0),
    Ads     = ifelse(ad == "Yes", 1, 0)
  ) %>%
  select(resp, task, choice, Netflix, Prime, Ads, price)

# Step 2: Sort for consistent grouping
mnl_data <- mnl_data %>%
  arrange(resp, task)

# Step 3: Confirm structure
str(mnl_data)
head(mnl_data)
```




## 4. Estimation via Maximum Likelihood

_todo: Code up the log-likelihood function._
```{r}
# Define the log-likelihood function for MNL
log_likelihood <- function(beta, data) {
  
  # Extract the design matrix X and outcome y
  X <- as.matrix(data[, c("Netflix", "Prime", "Ads", "price")])
  y <- data$choice
  
  # Create vector of linear utilities
  utility <- X %*% beta
  
  # Compute exponentiated utilities within choice sets
  data$exp_util <- exp(utility)
  
  # Compute denominators of the softmax (sum of exp utilities per task)
  data$denom <- ave(data$exp_util, data$resp, data$task, FUN = sum)
  
  # Compute choice probabilities
  data$prob <- data$exp_util / data$denom
  
  # Compute log-likelihood contribution per observation
  log_lik <- sum(log(data$prob[y == 1]))
  
  return(log_lik)
}
```


_todo: Use `optim()` in R or `scipy.optimize()` in Python to find the MLEs for the 4 parameters ($\beta_\text{netflix}$, $\beta_\text{prime}$, $\beta_\text{ads}$, $\beta_\text{price}$), as well as their standard errors (from the Hessian). For each parameter construct a 95% confidence interval._
```{r}
# Negative log-likelihood for minimization
neg_log_likelihood <- function(beta, data) {
  -log_likelihood(beta, data)
}

# Initial guess for beta coefficients
init_beta <- rep(0, 4)  # [Netflix, Prime, Ads, Price]

# Optimize using BFGS to get MLE
mle_result <- optim(
  par = init_beta,
  fn = neg_log_likelihood,
  data = mnl_data,
  method = "BFGS",
  hessian = TRUE
)

# Extract estimated betas
beta_hat <- mle_result$par

# Calculate standard errors from inverse Hessian
vcov_mat <- solve(mle_result$hessian)
se_beta <- sqrt(diag(vcov_mat))

# Construct 95% confidence intervals
z_value <- qnorm(0.975)
lower_95 <- beta_hat - z_value * se_beta
upper_95 <- beta_hat + z_value * se_beta

# Combine results into a data frame
results <- data.frame(
  Parameter = c("Netflix", "Prime", "Ads", "Price"),
  Estimate = round(beta_hat, 4),
  StdError = round(se_beta, 4),
  CI_Lower = round(lower_95, 4),
  CI_Upper = round(upper_95, 4)
)

# Print results
print(results)
```



## 5. Estimation via Bayesian Methods

_todo: code up a metropolis-hasting MCMC sampler of the posterior distribution. Take 11,000 steps and throw away the first 1,000, retaining the subsequent 10,000._

_hint: Use N(0,5) priors for the betas on the binary variables, and a N(0,1) prior for the price beta._

_hint: instead of calculating post=lik*prior, you can work in the log-space and calculate log-post = log-lik + log-prior (this should enable you to re-use your log-likelihood function from the MLE section just above)_

_hint: King Markov (in the video) use a candidate distribution of a coin flip to decide whether to move left or right among his islands.  Unlike King Markov, we have 4 dimensions (because we have 4 betas) and our dimensions are continuous.  So, use a multivariate normal distribution to pospose the next location for the algorithm to move to. I recommend a MNV(mu, Sigma) where mu=c(0,0,0,0) and sigma has diagonal values c(0.05, 0.05, 0.05, 0.005) and zeros on the off-diagonal.  Since this MVN has no covariances, you can sample each dimension independently (so 4 univariate normals instead of 1 multivariate normal), where the first 3 univariate normals are N(0,0.05) and the last one if N(0,0.005)._

```{r}
# Log-prior function
log_prior <- function(beta) {
  dnorm(beta[1], 0, 5, log=TRUE) +  # Netflix
    dnorm(beta[2], 0, 5, log=TRUE) +  # Prime
    dnorm(beta[3], 0, 5, log=TRUE) +  # Ads
    dnorm(beta[4], 0, 1, log=TRUE)    # Price
}

# Log-posterior function: log-likelihood + log-prior
log_posterior <- function(beta, data) {
  log_likelihood(beta, data) + log_prior(beta)
}

# MCMC settings
n_iter <- 11000
burn_in <- 1000
n_params <- 4
samples <- matrix(NA, nrow=n_iter, ncol=n_params)
colnames(samples) <- c("Netflix", "Prime", "Ads", "Price")

# Initialize chain at 0
current_beta <- rep(0, n_params)
current_log_post <- log_posterior(current_beta, mnl_data)

# Run Metropolis-Hastings MCMC
set.seed(123)
for (i in 1:n_iter) {
  
  # Proposal step: 4 independent normal proposals
  proposal <- current_beta + c(
    rnorm(1, 0, sqrt(0.05)),
    rnorm(1, 0, sqrt(0.05)),
    rnorm(1, 0, sqrt(0.05)),
    rnorm(1, 0, sqrt(0.005))
  )
  
  # Calculate log-posterior at proposal
  proposal_log_post <- log_posterior(proposal, mnl_data)
  
  # Acceptance probability
  log_accept_ratio <- proposal_log_post - current_log_post
  if (log(runif(1)) < log_accept_ratio) {
    current_beta <- proposal
    current_log_post <- proposal_log_post
  }
  
  # Store current draw
  samples[i, ] <- current_beta
}

# Discard burn-in
posterior_samples <- samples[(burn_in+1):n_iter, ]

# Summarize posterior
posterior_summary <- apply(posterior_samples, 2, function(x) {
  c(Mean=mean(x), SD=sd(x), 
    CI_Lower=quantile(x, 0.025),
    CI_Upper=quantile(x, 0.975))
})
round(t(posterior_summary), 4)
```



_todo: for at least one of the 4 parameters, show the trace plot of the algorithm, as well as the histogram of the posterior distribution._
```{r}
# Plot trace and histogram for β_price (column 4)
par(mfrow=c(1,2))  # two plots side by side

# Trace plot
plot(posterior_samples[, 4], type="l", col="blue",
     main=expression("Trace plot: " * beta[price]),
     xlab="Iteration", ylab=expression(beta[price]))

# Histogram
hist(posterior_samples[, 4], breaks=40, col="lightblue",
     main=expression("Posterior of " * beta[price]),
     xlab=expression(beta[price]), freq=FALSE)

# Add kernel density
lines(density(posterior_samples[, 4]), col="darkblue", lwd=2)
```


_todo: report the 4 posterior means, standard deviations, and 95% credible intervals and compare them to your results from the Maximum Likelihood approach._

```{r}
# Rebuild the MLE summary table
results <- data.frame(
  Parameter = c("β_Netflix", "β_Prime", "β_Ads", "β_Price"),
  Estimate = round(mle_result$par, 4),
  StdError = round(sqrt(diag(solve(mle_result$hessian))), 4),
  CI_Lower = round(mle_result$par - qnorm(0.975) * sqrt(diag(solve(mle_result$hessian))), 4),
  CI_Upper = round(mle_result$par + qnorm(0.975) * sqrt(diag(solve(mle_result$hessian))), 4)
)
```

```{r}
# Compute posterior summaries
posterior_summary <- apply(posterior_samples, 2, function(x) {
  c(Mean = mean(x), 
    SD = sd(x), 
    CI_Lower = quantile(x, 0.025), 
    CI_Upper = quantile(x, 0.975))
})
posterior_summary <- round(t(posterior_summary), 4)

# Set column and row names
colnames(posterior_summary) <- c("Mean", "SD", "CI_Lower", "CI_Upper")
rownames(posterior_summary) <- c("β_Netflix", "β_Prime", "β_Ads", "β_Price")
```

```{r}
# Create long-format MLE and Bayesian tables
mle_df <- data.frame(
  Parameter = results$Parameter,
  Method = "MLE",
  Estimate = results$Estimate,
  Std_Error = results$StdError,
  CI_Lower = results$CI_Lower,
  CI_Upper = results$CI_Upper
)

bayesian_df <- data.frame(
  Parameter = rownames(posterior_summary),
  Method = "Bayesian",
  Estimate = posterior_summary[, "Mean"],
  Std_Error = posterior_summary[, "SD"],
  CI_Lower = posterior_summary[, "CI_Lower"],
  CI_Upper = posterior_summary[, "CI_Upper"]
)

# Combine both
long_comparison <- rbind(mle_df, bayesian_df)

rownames(long_comparison) <- NULL

# Display final table
knitr::kable(long_comparison, 
             caption = "Comparison of MLE and Bayesian Posterior Summaries",
             align = "l")
```







## 6. Discussion

_todo: Suppose you did not simulate the data. What do you observe about the parameter estimates? What does $\beta_\text{Netflix} > \beta_\text{Prime}$ mean? Does it make sense that $\beta_\text{price}$ is negative?_

1. General Observations About the Estimates

If you had not simulated the data and instead received it from a real conjoint survey:

The estimated coefficients from both MLE and Bayesian MCMC represent average marginal utilities (part-worths) associated with each attribute level across respondents.
These values reflect relative preferences, not absolute value. What matters is their magnitude and sign.
2. Interpretation of $\beta_\text{Netflix} > \beta_\text{Prime}$

This means:

Consumers, on average, derive more utility (i.e., prefer more) from Netflix than Amazon Prime, holding all else constant (ads and price).
Since Hulu is the baseline (reference category), both coefficients represent preferences relative to Hulu. So:
$\beta_\text{Netflix} > 0$ implies Netflix is preferred over Hulu.
$\beta_\text{Prime} > 0$ implies Prime is also preferred over Hulu.
But $\beta_\text{Netflix} > \beta_\text{Prime}$ implies Netflix is preferred over Prime as well.
This is consistent with real-world brand strength if users generally perceive Netflix as having superior content or experience.

3. Does It Make Sense That $\beta_\text{price} < 0$?

Yes — absolutely.

A negative coefficient on price indicates that higher prices reduce utility, making a product less likely to be chosen, which matches rational consumer behavior.
The magnitude reflects price sensitivity. A steeper negative value would imply that price is a more dominant decision factor for respondents.
✅ Bottom Line:
Even without simulation, the estimated values:

Make intuitive and economic sense.
Suggest a strong brand preference for Netflix, moderate for Prime, and cost-sensitivity in choice behavior.
The signs and relative magnitudes align well with what we'd expect from actual consumer preferences.

_todo: At a high level, discuss what change you would need to make in order to simulate data from --- and estimate the parameters of --- a multi-level (aka random-parameter or hierarchical) model. This is the model we use to analyze "real world" conjoint data._

To simulate and estimate a multi-level (also known as random-parameter or hierarchical) multinomial logit model, we would need to move from a single set of utility coefficients shared across all respondents to individual-specific preference parameters. This means each respondent $i$ would have their own vector of part-worth utilities $\beta_i$, drawn from a population distribution, typically assumed to be multivariate normal: $\beta_i \sim \mathcal{N}(\mu, \Sigma)$. Simulating such data would require generating $\beta_i$ values for each individual and using them to compute utility and choices. Estimating this model is more complex—it involves recovering both the individual-level parameters $\beta_i$ and the population-level hyperparameters $\mu$ and $\Sigma$. This is usually done through Bayesian MCMC methods (such as Gibbs sampling or Hamiltonian Monte Carlo) or simulated maximum likelihood approaches. Multi-level models are more computationally intensive but allow for richer and more realistic representations of individual preference heterogeneity in real-world conjoint data.












